An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-01-10 Paper/BoundEffRed.tex
review

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR). There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, the low computational time allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many TF representations. An additional application to ECG is included in the Supplementary Materials.

>>>>

Attempts to minimize the boundary effects generally consists in softening the discontinuity on signal edges. Usually, the approaches can be classified into two main classes, choosing a proper analysis window and extending/forecasting the signal. In the first class, judicious choice of analysis window whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane, for instance, \cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship between the chosen analysis windows and the TF analysis tool, these techniques do not make it possible to reduce the boundary effects of all TF representations. 
%
Another natural idea, the second class, consists of carrying out a preliminary step of extending, or forecasting, the signal beyond its boundaries. Due to its flexibility, various forecasting schemes have been proposed. For example, there exist simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. 
%
There exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD) and the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR).
% 
{\color{red}
In speech processing, dynamic mode predictors have also been proposed~\cite{Vargas11speech} to forecast signals falling into the so-called \textit{source--filter} model. This gradient-based technique rely on the shadowing approach proposed in~\cite{Grebogi90shadowing}.
} 
%
There also exist extension schemes based on stochastic models, such as the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}. 
%
In the physically relevant dynamical models, the oscillation or trend, are usually modeled as the mean of a stochastic process, while in the stochastic models, the oscillation or trend are modeled by the covariance structure of the stochastic process. 
It is also possible to consider polynomial regression~\cite{fan1996local} or modeling the mean by splines~\cite{hall2005theory}, kernel functions~\cite{chang2010training}, and wavelets~\cite{marron1998exact}, or nonparametric regression~\cite{fan1996local}, to estimate the mean function before forecasting the signal. Neural network models, such as the long short-term memory model~\cite{vlachas2018data}, is another approach. 
%
The above list is far from exhaustive, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. 
%
While the model-based extension scheme gives better-extended signals than those simple extension without dynamical models, they generally have a great computational cost.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the real-time reduction of boundary effects in TF representations. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient to extend signals formed by a sum of sine waves. Moreover, the low computational time allows us to switch to a real-time implementation of {\sf BoundEffRed}, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many TF representations. An additional applications to ECG is included in the Supplementary Materials.

### 2021-01-10 Paper/algorithm.tex
review

\paragraph{Dynamical Model} 
Establishing a dynamical model consists of determining the relation linking $\bY$ to $\bX$, that is finding a function $f$ so that
\begin{equation}
\bY = f(\bX)\ .
\label{eq:generic.model}
\end{equation}
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see, in section~\ref{se:theoretical}, that considering such a complex dynamic model is not necessary for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization of a nonlinear system}, when the system is sufficiently regular. Notice that this linearized dynamical model can be written equivalently according to the subsignals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}
From the Takens' embedding perspective~\cite{Takens81detecting}, $\bX$ can be viewed as a lag map. Under the manifold assumption of the intrinsic phase space that hosts the dynamics, $f$ describes the dynamics supported on the manifold space, and $\bA$ approximates the dynamics on the manifold.

\paragraph{Forecasting}
The forecasting method consists in estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. {\color{red}We then obtain the forecasting of $\bx_{k+\ell}$ by recursively applying the linear relation~\eqref{eq:dyn.model}:
\begin{equation*}
\tilde\bx_{K+\ell} = \underbrace{\tilde\bA\tilde\bA\cdots\tilde\bA}_{\scriptstyle\ell\mbox{ \footnotesize {times}}}\bx_{K} = \tilde\bA^\ell\bx_{K}\ .
\end{equation*}  
Let $\be_M$ denote the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$. Consequently, given definition~\eqref{eq:xk}, the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ is provided by:
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \be_M^T\tilde\bx_{K+\ell} = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, \ie~$\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell$.
}

>>>>

\paragraph{Dynamical Model} 
Establishing a dynamical model consists of determining the relation linking $\bY$ to $\bX$, that is finding a function $f$ so that
\begin{equation}
\bY = f(\bX)\ .
\label{eq:generic.model}
\end{equation}
In a general framework, forecasting means estimating the function $f$ from the observed values taken by the signal, in order to predict its future values. For instance, the dynamic mode decomposition~\cite{Schmid10dynamic,Williams15data} or other more complicated models \cite{Roberts13Gaussian,DeLivera11forecasting,west2006bayesian,vlachas2018data}, allow this by setting additional constraints on the behavior of $f$. We will see, in section~\ref{se:theoretical}, that considering such a complex dynamic model is not necessary for the study of the oscillatory signals of interest to us. That is why we consider here a naive dynamical model, assuming that we have the following relation:
\begin{equation}
\bY = \bA\bX\ ,
\label{eq:dyn.model}
\end{equation}
where $\bA\in\RR^{M\times M}$. In other words, we adopt a classical strategy in the study of dynamical systems, the {\em linearization of a nonlinear system}, when the system is sufficiently regular. Notice that this linearized dynamical model can be written equivalently according to the subsignals $\bx_k$, as:
\begin{equation}
\bx_{k+1} = \bA\bx_{k},\ ,\forall k\in\{0,\dots,K-1\}\ .
\end{equation}
%
{\color{blue}
From the Takens' embedding perspective~\cite{Takens81detecting}, $\bX$ can be viewed as a lag map. Under the manifold assumption of the intrinsic phase space that hosts the dynamics, $f$ describes the dynamics supported on the manifold space, and $\bA$ approximates the dynamics on the manifold.
}

\paragraph{Forecasting}
The forecasting method consists in estimating the unknown matrix $\bA$. Indeed, let $\tilde\bA$ denotes the estimate of $\bA$. 
%
{\color{red}
We then obtain the forecasting of $\bx_{k+\ell}$ by recursively applying the linear relation~\eqref{eq:dyn.model}:
\begin{equation*}
\tilde\bx_{K+\ell} = \underbrace{\tilde\bA\tilde\bA\cdots\tilde\bA}_{\scriptstyle\ell\mbox{ \footnotesize {times}}}\bx_{K} = \tilde\bA^\ell\bx_{K}\ .
\end{equation*}  
Let $\be_M$ denote the unit vector of length $M$ given by $\be_M = \begin{pmatrix} 0 & \cdots & 0 & 1\end{pmatrix}^T$. Consequently, given definition~\eqref{eq:xk}, the forecasting of the signal at time $\frac{N-1+\ell}{\fs}$ is provided by
\begin{equation}
\label{eq:prediction}
\tilde \bx [N-1+\ell] = \be_M^T\tilde\bx_{K+\ell} = \balpha^{(\ell)}\bx_{K}\ ,
\end{equation}  
where $\balpha^{(\ell)}$ denotes the last row of $\tilde\bA^\ell$, \ie~$\balpha^{(\ell)}=\be_M^T\tilde\bA^\ell$.
}

### 2021-01-10 Paper/results.tex
review

As anticipated by Theorem~\ref{th:error}, the variance of the estimator decreases linearly with $\frac1K$ as long as the dataset size is large enough. Indeed, in the log-log plot displayed in Fig.~\ref{fig:res.size.sine}, as soon as $K>700$, each of the three curves representing the forecasting variance as a function of $K$ tends towards an asymptote of slope $-1$.

@@

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results, in particular in comparison with the point-symmetric extension. Besides, even though the other more sophisticated methods, like GPR, give MSE values that have a slightly smaller standard deviation, these methods are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time. {\color{red} Besides, the last column of Table~\ref{tab:THO} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.}

\begin{table}
\centering
\caption{AHM Signal: Performance of the Extension Methods and the Associated Boundary-Free STFT}
\begin{tabular}{|c||c|c||c|}
  \hline
   \multirow{2}{60pt}{\centering Extension method}  & \multirow{2}{30pt}{\centering CPU time (s)} & \multirow{2}{50pt}{\centering MSE\\ (mean $\pm$ SD)} & \multirow{2}{50pt}{\centering Index $D$ of the STFT}\\
    &  & & \\
   \hhline{|=#=|=#=|}
   {\sf SigExt ($M=100$)} & $0.008$ & $1.133 \pm  0.077$ & \\
   \hline
   {\sf SigExt ($M=750$)} & $0.122$ & $0.479 \pm  0.166$ & \\
   \hline
   {\sf SigExt ($M=1500$)} & $0.634$ & $0.907 \pm  0.377$ & \\
   \hline
   {\sf SigExt ($M=3000$)} & $2.881$ & $5.365 \pm  3.919$ & \\
   \hline
   Symmetric &  {\color{red}$<0.001$} & {\color{red}$4.368 \pm  0.001$} &\\
   \hline
   EDMD & {\color{red}$11.10$} & {\color{red}$1.156  \pm  0.003$} & \\
   \hline
   GPR & {\color{red}$244.7$} & {\color{red}$1.025  \pm  0.403$} &\\
   \hline
   TBATS  & {\color{red}$633.6$} & {\color{red}$1.501 \pm  1.354$} & \\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

\subsection{Real Physiological Signals}

>>>>

As anticipated by Theorem~\ref{th:error}, the variance of the estimator decreases linearly with $\frac1K$ as long as the dataset size is large enough. Indeed, in the log-log plot displayed in Fig.~\ref{fig:res.size.sine}, as soon as $K>700$, each of the three curves representing the forecasting variance as a function of $K$ tends towards an asymptote of slope $-1$. 
%
{\color{red}
In practice, the asymptotic behavior of {\sf SigExt}, illustrated here on a synthetic signal, can be generalized to the study of real-life signals, as long as the duration of the recorded signal largely exceeds that of the desired extension, which allows the user to set $K\gg L$.
}

@@

\paragraph{Comparison of the extensions methods}
The same set of experiments is run on the extension methods outlined in section~\ref{sse:methods}. The results, displayed in Table~\ref{tab:mse.sine}, show that the naive extension we propose gives satisfying results, in particular in comparison with the point-symmetric extension. Besides, even though the other more sophisticated methods, like GPR, give MSE values that have a slightly smaller standard deviation, these methods are substantially limited by the computing time they require, which prevents them from being used to exploit real-time data. Thus, {\sf SigExt} is the extension method that optimizes the trade-off between forecasting quality and computing time.
%
{\color{red}
Besides, the last column of Table~\ref{tab:mse.sine} gives the performance index $D$ of the boundary-free STFT associated with each extension method. This illustrates the ability of our algorithm to reduce boundary effects to a level comparable to what is possible with the GPR or TBATS extensions. The performance of {\sf BoundEffRed} on other TF representations is more thoroughly analyzed on real-life signals in the following.
}

\begin{table}
\centering
\caption{AHM Signal: Performance of the Extension Methods and the Associated Boundary-Free STFT}
\begin{tabular}{|c||c|c||c|}
  \hline
   \multirow{2}{60pt}{\centering Extension method}  & \multirow{2}{30pt}{\centering CPU time (s)} & \multirow{2}{50pt}{\centering MSE\\ (mean $\pm$ SD)} & \multirow{2}{50pt}{\centering Index $D$ of the STFT}\\
    &  & & \\
   \hhline{|=#=|=#=|}
   {\sf SigExt ($M=100$)} & $0.008$ & $1.133 \pm  0.077$ & {\color{blue}$x \pm  x$}\\
   \hline
   {\sf SigExt ($M=750$)} & $0.122$ & $0.479 \pm  0.166$ & {\color{blue}$x \pm  x$}\\
   \hline
   {\sf SigExt ($M=1500$)} & $0.634$ & $0.907 \pm  0.377$ & {\color{blue}$x \pm  x$}\\
   \hline
   {\sf SigExt ($M=3000$)} & $2.881$ & $5.365 \pm  3.919$ & {\color{blue}$x \pm  x$}\\
   \hline
   Symmetric &  {\color{blue}$<0.001$} & {\color{blue}$x \pm  x$} & {\color{blue}$x \pm  x$}\\
   \hline
   EDMD & {\color{blue}$x$} & {\color{blue}$x  \pm  x$} & {\color{blue}$x \pm  x$}\\
   \hline
   GPR & {\color{blue}$x$} & {\color{blue}$x  \pm  x$} & {\color{blue}$x \pm  x$}\\
   \hline
   TBATS  & {\color{blue}$x$} & {\color{blue}$x \pm  x$} & {\color{blue}$x \pm  x$}\\
   \hline
\end{tabular}
\label{tab:mse.sine}
\end{table}

\subsection{Real Physiological Signals}
\label{sse:physio.sig}
